---
title: "Lab 4: Factors"
author: "Environmental Data Analytics | John Fay and Luana Lima "
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---


```{r setup}
knitr::opts_chunk$set(message = FALSE, warning = FALSE) #ignore for now?
#Read packages
library(tidyverse);library(lubridate); library(here)

#Read in data, don't convert strings to factors
sites <- read.csv(here('Data/Raw/NWIS_SiteFlowData_NE_RAW.csv')) #no strings as factors = true piece that we'd had on prior sites, bc the variables should be int, num, etc)

dim(sites)

#Convert date column to date object  #lubridate don't need to learn the format... cheat sheet below
sites$date <- ymd_hms(sites$date)

```


## Examine the lakename column as a character column
* Tip - Lubridate cheatsheet: <https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf>
```{r create and explore days of week column}

#Compute the day of the week each sample was collected #dow: day of the week, wday: returns day of the week, sunday-1, mon-2, etc.
sites$dow <- wday(sites$date)

## What class is the dow column?
class(sites$dow)

#List the unique values in the column #not all numbers have numeric meaning... cant average days of week
unique(sites$dow)

#Summary - is it meaningful? #reporting as categories, rather than numbers (1, 148, 3)
summary(sites$dow)

```

## Convert the days of the week to a factor
```{r convert dow column to factor and explore}

#Convert to a factor --- categories, grouping variable, most of the time strings are factors except ie: dna sequence, where the values mean stuff

sites$dow <- factor(sites$dow)

## What class is the dow column now?
class(sites$dow)

#List the unique values in the column
unique(sites$dow)

#Summary - is it meaningful?
summary(sites$dow)

#Show the levels associated with our factor column
levels(sites$dow)
```

## Factors & Levels
A **factor** is a grouping variable: rows with the same value in the factor column are members of the same group. For example, all rows with a `dow` of 3 were collected on a Tuesday.

**Levels** are the unique values that the factors can take. The `unique()` function reveals the levels as does the `levels()` function. By default, the levels includes only the values in the factor column. However, we can add additional levels and we can also re-order levels. Why would we want to do this? Well, one example would be if we wanted to plot the number of samples recorded each day of the week, and in that plot we wanted to include all days of the week, not just the one's found in our data...

```{r plot samples recorded each day of the week}
#Create a bar plot showing the count of samples by day of week, bar plot makes sense bc categorical
ggplot(sites,aes(x=dow)) +
  geom_bar()
```



```{r add levels to our data}
#Update the levels to include all seven days of the week (note we have to start with existing levels)
levels(sites$dow) <- c(levels(sites$dow),"1","2","6","7")

#Confirm the change
levels(sites$dow)

#Plot again
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)

```

We now have all days of the week, but the order is incorrect. Also, we have day numbers, not day names. How to fix?
```{r reorder and rename levels}

#Reorder by overwriting the column with the existing values assigned with a new order of levels
sites$dow <- factor(sites$dow, levels=c("1","2","3","4","5","6","7")) #Or... as.character(seq(1,7))

#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)

#Add **labels** to our factors
sites$dow <- factor(
  sites$dow, #original values
  levels=c("1","2","3","4","5","6","7"), #here are given levels
  labels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat")) #here are provided labels

#See that the levels are updated
levels(sites$dow)

#View the plot
ggplot(sites,aes(x=dow)) +
  geom_bar() + 
  scale_x_discrete(drop=FALSE)
```

## An exercise for you: see if you can plot the number of samples recorded each month