---
title: "SamanthaSedar_A02_Coding Basics"
author: "Samantha Sedar"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. 
seq(1, 30, 3) #sequence- from, to, by#
one_to_thirty_by_3<-seq(1, 30, 3) #naming

#2. 
mean(one_to_thirty_by_3) #calculate mean
median(one_to_thirty_by_3) #calculate median

#3. #determine that they are equal 
mean(one_to_thirty_by_3)<=median(one_to_thirty_by_3) 
mean(one_to_thirty_by_3)>=median(one_to_thirty_by_3)
```

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

#5 and #6
names<-c("Jack", "Jill", "John", "Jane") #values/string
scores<-c(75, 100, 85, 50) #numbers
passed<-c(TRUE, TRUE, TRUE, FALSE) #boolean



#7

df_names <- as.data.frame(names)

df_students_name_score_pass <- cbind(names, scores, passed)

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer:
This data frame consists of four different components, the students' names, scores, and whether or not they passed the test. A matrix would only consist of one class of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
#10
x <- scores
passing_grade <- function(x){
    ifelse((x>=50), print("TRUE"),   print("FALSE")) 
}


#11
passing_grade(x)


```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 'ifelse' worked because passing is applicable for the entire vector, rather than each score being tested on distinct parameters.
